\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\setlength{\parskip}{0.8em}
\setlength{\parindent}{0pt}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true, linkcolor=blue!60!black, citecolor=blue!60!black, urlcolor=blue!60!black]{hyperref}
\usepackage{fontspec}
\usepackage{xeCJK}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{polyglossia}
\setdefaultlanguage{english}
\newfontfamily\greekfont{Libertinus Serif}[Script=Greek]
\setmainfont{Libertinus Serif}
\captionsetup[figure]{labelfont=bf, font=small, margin=10pt}
\usepackage{enumitem}
\setlist[itemize]{left=0pt, label=--, itemsep=0.5em}
\setlist[enumerate]{left=0pt, itemsep=0.5em}
\newtheorem{concept}{Concept}
\newcommand{\symtry}{\mathbin{/}}
\newcommand{\goldenset}{\varnothing}
\newcommand{\knotinfinity}{\textnormal{0}}
\title{\Huge\bfseries Axiomatic Framework Integrating NBG, TG, and TOI}
\author{Converted from\texttt{ draft2.txt }}
\date{\today}
\begin{document}
\maketitle
Great. I will construct a formal axiomatic framework that integrates NBG as the foundational language, TG axioms to ensure the existence of a hierarchy of Grothendieck universes, and TOI’s principles of absolute infinity, symmetry, and nested contexts.

This framework will support formal mathematics and be grounded in a Philosophy of Science perspective, emphasizing explanatory clarity and coherence with scientific modeling. I’ll structure the output as a formal system: including primitive notions, axioms, and a brief rationale for each axiom class.

I'll begin this synthesis and let you know once the full architectonic foundation is ready for review.


\section*{Axiomatic Framework Integrating NBG, Tarski–Grothendieck, and Theory of Infinity}

\subsection*{Primitive Notions and Language}

We work in a first-order logic framework with **two types of entities**: **sets** and **classes**. Every set is a class, but **proper classes** are collections “too large” to be sets (they are not members of any other class). The language includes:

\begin{itemize}
  \item The binary **membership** relation \$\in\$ (with \$x \in A\$ meaning set \$x\$ is an element of class \$A\$).
  \item A unary predicate **\$\mathsf{Set}(x)\$** to assert that a class \$x\$ is actually a set. (If \$\neg\mathsf{Set}(X)\$, then \$X\$ is a proper class.)
  \item **Constant symbols** for special classes: **\$\infty\$** (the “Absolute Infinite”) and **\$\Sym\$** (a symmetry group acting on \$\infty\$). We will define **Golden Sets** via the action of \$\Sym\$.

\end{itemize}
**Definition – Proper Class**: A class \$C\$ is *proper* (not a set) if it cannot be an element of any class. Equivalently, \$C\$ is a set if and only if there exists some class \$B\$ with \$C \in B\$. For example, \$\infty\$ (the class of **all** sets) is a proper class – if \$\infty\$ were a set, it would lead to the Russell paradox, so \$\infty\$ must not belong to any class.

We now present the axioms of the system. They are grouped into: (a) **NBG set theory axioms** (governing sets, classes, and ZFC-like set construction), (b) the **Tarski–Grothendieck (TG) axiom** ensuring large infinities (Grothendieck universes), and (c) additional **Theory of Infinity (TOI) axioms** introducing \$\infty\$, the symmetry group \$\Sym\$, and Golden Sets (nested invariant contexts). Each axiom is stated formally (in the language of NBG) and followed by an explanation of its role in mathematics and the philosophy of science (e.g. how it contributes to explanatory adequacy, structural clarity, or modeling capacity).

\subsection*{Axioms of NBG Set Theory (Base Formalism)}

\begin{enumerate}
  \item **Extensionality:** \$\forall x,\forall y,\big\[,\forall z,(z \in x \Leftrightarrow z \in y)\ \rightarrow\ x = y,\big]\$.
\end{enumerate}
   *Explanation:* If two classes (in particular, two sets) have exactly the same elements, then they are identical. This standard axiom guarantees well-defined identity of collections – a foundational principle for any set theory. It aligns with **structural clarity**: any mathematical object is determined by its constituent elements, nothing more.

\begin{enumerate}
  \item **Class Comprehension (Schema):** For any formula \$\phi(x)\$ *that quantifies only over sets*, there exists a class \$C = {,x : \phi(x),}\$ such that \$\forall x,\[x \in C \Leftrightarrow \phi(x)]\$.
\end{enumerate}
   *Explanation:* This axiom (really a schema) lets us form classes by separating sets that satisfy a property. It ensures **expressiveness**: we can define collections like “the class of all sets with property P” without restriction – even if that class is too large to be a set. This is crucial for handling “inconsistent multiplicities” (Cantor’s term) as classes rather than sets, thereby **avoiding paradoxes** while still talking about totalities like “the class of all sets” or “the class of all ordinals”.

\begin{enumerate}
  \item **Empty Set:** \$\exists x,\forall y,(y \notin x)\$.
\end{enumerate}
   *Explanation:* There is a set with no members, denoted \$\varnothing\$. The Empty Set axiom guarantees a **starting point** for building the universe of sets. In philosophical terms, it posits a null object – a baseline context from which all other mathematical objects can be constructed.

\begin{enumerate}
  \item **Pairing:** \$\forall a,\forall b,\exists p,\forall x,\[x \in p \Leftrightarrow (x=a \vee x=b)]\$.
\end{enumerate}
   *Explanation:* For any two sets \$a,b\$, there is a set \$p={a,b}\$ containing exactly those two. This allows construction of finite sets and ensures that even simple combinations of objects exist as sets. It contributes to **explanatory adequacy** by enabling the discrete bundling of entities (e.g. forming a system composed of two parts).

\begin{enumerate}
  \item **Union:** \$\forall A,\exists U,\forall x,\[x \in U \Leftrightarrow \exists y,(y \in A \wedge x \in y)]\$.
\end{enumerate}
   *Explanation:* For any set \$A\$ of sets, there is a set \$U=\bigcup A\$ that unites all elements of elements of \$A\$. This axiom allows building larger contexts from smaller ones (e.g. forming a grand system from subsystems). It reflects the principle that we can **merge contexts** into a bigger context when modeling phenomena (useful in science when combining systems).

\begin{enumerate}
  \item **Power Set:** \$\forall A,\exists P,\forall x,\[x \in P \Leftrightarrow x \subseteq A]\$.
\end{enumerate}
   *Explanation:* For any set \$A\$, there is a set \$P=\mathcal{P}(A)\$ of all subsets of \$A\$. Power sets are essential for constructing function spaces, probability spaces, etc. In foundational terms, this axiom ensures that the universe is **rich in structure** (given any context \$A\$, one can consider all subcontexts). It contributes to **structural clarity** by guaranteeing that every definable sub-collection of a set is itself a set (within our framework, large sub-collections might be proper classes, but any subset of a set *is* a set).

\begin{enumerate}
  \item **Infinity:** There exists a set \$I\$ such that \$\varnothing \in I\$ and \$\forall x,(x \in I \rightarrow x \cup {x} \in I)\$.
\end{enumerate}
   *Explanation:* This axiom asserts the existence of an infinite set (one that contains the empty set, \${ \emptyset }\$, \${\emptyset,{\emptyset}}\$, and so on). It seeds the **iterative hierarchy** with an infinite structure (often taken to model the natural numbers). In scientific modeling, this allows us to presume the existence of **countably infinite** structures (like an infinite sequence of events or states), which is often a minimal requirement for mathematics (and physics, e.g. time series). (We will see that even larger infinities are guaranteed by the TG axiom below.)

\begin{enumerate}
  \item **Replacement (Schema):** If \$F\$ is a definable class function and \$X\$ is a set, then \$F\[X] = {F(x) : x \in X}\$ is also a set.
\end{enumerate}
   *Explanation:* This schema captures the idea that the image of any set under a definable transformation is still a set. It ensures the universe is closed under definable mappings, which is vital for building complicated objects without leaving the realm of sets. Replacement contributes to **explanatory adequacy** by allowing the construction of new sets via rule-based transformations of existing sets (mirroring how one can obtain new states of a system from given states by applying rules).

\begin{enumerate}
  \item **Foundation (Regularity):** \$\forall x,\[x \neq \varnothing \rightarrow \exists y \in x:; x \cap y = \varnothing]\$.
\end{enumerate}
   *Explanation:* Every non-empty set \$x\$ has some member \$y\$ that shares no element with \$x\$ (in particular \$y \notin x\$). This prevents any infinite descending \$\in\$-chains (no \$x \ni x\_1 \ni x\_2 \ni \cdots\$) and rules out loops like \$x \in x\$. Foundation secures a **well-founded** universe where sets are built “from the ground up” (empty set, singletons, pairs, etc., cumulatively). Philosophically, this axiom enforces a baseline for **structural clarity** – every element of a set comes from a strictly “lower” rank, preventing pathological self-reference.

\begin{enumerate}
  \item **Global Choice (optional):** There exists a class function \$G\$ on the class of all non-empty sets such that for every non-empty set \$A\$, \$G(A) \in A\$.
\end{enumerate}
    *Explanation:* This is a strong form of the Axiom of Choice at the class level. It asserts a uniform way to pick an element from every set. While not strictly required for the framework, assuming Global Choice simplifies reasoning about large collections and is often adopted in NBG. In scientific modeling, a choice function can represent a consistent selection of states or representatives from equivalence classes, aiding **constructive clarity** (though the axiom is more technical than philosophical).

> **Note:** Axioms 3–10 above correspond to the usual Zermelo–Fraenkel set axioms (adapted to classes when needed). They ensure that within our framework, all the standard set constructions and properties of ZFC hold. **NBG is a conservative extension of ZF**, meaning any theorem about sets provable here is already provable in ZF. The class apparatus is used for convenience (finite axiomatization and talking about “big” collections like \$\infty\$ itself).

\subsection*{Tarski–Grothendieck Axiom (Large Universes)}

\begin{enumerate}
  \item **Grothendieck Universe Axiom:** *For every set, there exists a larger set that is a “universe” containing it.* Formally, \$\forall x,\exists U,\Big\[\mathsf{Set}(U)\ \wedge\ x \in U\ \wedge\ \text{Universe}(U)\Big]\$. Here **Universe\$(U)\$** means:
\end{enumerate}

\begin{itemize}
  \item (Transitivity) \$\forall a,\forall b,\[,a \in b \wedge b \in U ;\rightarrow; a \in U,]\$.
  \item (Closure under Pairing) \$\forall a,!b\in U:\ {a,b} \in U\$.
  \item (Closure under Power Set) \$\forall a\in U:\ \mathcal{P}(a) \in U\$.
  \item (Closure under Union) \$\forall A\in U:\ \bigcup A \in U\$.
  \item (Optionally, Closure under Replacement as well, so that \$U\$ is an elementary submodel of the whole universe.)

\end{itemize}
    *Explanation:* This axiom asserts an **inexhaustible hierarchy of infinities**. Every set \$x\$ lies inside some larger set \$U\$ that is *transitively closed* and closed under basic set operations – in other words, \$U\$ is a **Grothendieck universe** (a model of ZF set theory in its own right). Such a \$U\$ can be thought of as a *self-contained mathematical cosmos* encompassing \$x\$. This axiom guarantees the existence of **strongly inaccessible cardinals** – in fact, it implies there are *infinitely many, unboundedly large* inaccessible cardinals (each universe \$U\$ is roughly \$V\_\kappa\$ for some inaccessible \$\kappa\$).

    *Rationale:* In pure mathematics, Grothendieck introduced universes to **avoid proper classes** in category theory – for example, one can treat “the class of all sets of moderate size” as a set \$U\$, so that the “category of all sets in \$U\$” is legitimately a set-level entity inside an even larger universe. This adds **structural clarity**: it becomes possible to talk about “categories of all sets” or “the power set of a proper class” by relocating the discussion to an appropriate universe \$U\$. In our axiomatic framework, TG ensures a plenitude of infinities for **explanatory adequacy** – we are never stuck at a largest set, since for any system we have, a bigger context exists containing it. This is harmonious with scientific modeling needs: it means one can always embed a model into a larger domain if needed (reflecting the intuition that any given theory can be encompassed by a more comprehensive theory). Moreover, each Universe \$U\$ is a *closed arena* for mathematics, supporting internal reasoning as if \$U\$ were the entire universe of sets. This aligns with the **reflection principle** (originally noted by Cantor): any property of the Absolute Infinite (the total universe) is mirrored in some smaller realm. Thus, TG axioms enhance the **explanatory power** of the framework by ensuring that structures of any conceivable size are available, while maintaining consistency through the distinction of sets vs. classes.

\subsection*{Theory of Infinity (TOI) Axioms}

\begin{enumerate}
  \item **Absolute Infinite Class (\$\infty\$):** \$\infty\$ is a distinguished proper class that contains every set: \$\forall x,\[\mathsf{Set}(x) \rightarrow x \in \infty]\$. Moreover, \$\neg\mathsf{Set}(\infty)\$ (so \$\infty\$ is not itself a set).
\end{enumerate}
    *Explanation:* \$\infty\$ axiomatizes the idea of **Cantor’s Absolute Infinite** – an all-encompassing infinity that cannot be a set without contradiction. In NBG terms, \$\infty\$ is essentially the class of all sets (often denoted \$V\$ in set theory). The axiom says every set resides in this universal class, but \$\infty\$ has “too many elements” to be a set. This provides a **single, unified backdrop** for the hierarchy of all sets. Philosophically, introducing \$\infty\$ as a proper class acknowledges that the totality of mathematical objects is an entity we can reason about, but only via **higher-order** logic (as a class) rather than as a set. This move preserves consistency while allowing discourse about “the universe as a whole.” It aligns with the Philosophy of Science by making explicit the **ultimate context** for any mathematical or scientific structure: any specific model or system (set) is situated in the Absolute Infinite class \$\infty\$, analogous to viewing a scientific phenomenon as part of the entire cosmos.

\begin{enumerate}
  \item **Symmetry Group on \$\infty\$ (Sym):** There exists a class \$\Sym\$ and a binary operation \$\circ\$ on \$\Sym\$ (functional composition) such that \$(\Sym,\circ)\$ is a group of transformations acting on \$\infty\$. Formally, each \$g \in \Sym\$ is (interpreted as) a bijection \$g: \infty \to \infty\$ (permutation of the universe), and:
\end{enumerate}

\begin{itemize}
  \item (Closure) \$\forall f,g \in \Sym:\ f\circ g \in \Sym\$.
  \item (Identity) \$\exists e \in \Sym:\ \forall x\in \infty,\ e(x)=x\$.
  \item (Inverses) \$\forall g \in \Sym:\ \exists g^{-1}\in \Sym:\ \forall x,\ g^{-1}(g(x))=x\$.
  \item (Action on \$\infty\$) \$\forall g\in\Sym,\ \forall x\in \infty:\ g(x)\in \infty\$ (guaranteed since \$g\$ is a permutation of \$\infty\$).

\end{itemize}
    *Explanation:* This axiom posits a notion of **symmetry at the level of the entire universe of sets**. Intuitively, \$\Sym\$ is the class of all automorphisms of the universe \$(\infty,\in)\$, or at least some large group of bijections of \$\infty\$ that we assume exists. (In practice, to avoid paradox, \$\Sym\$ may be restricted to *definable* or *constructible* permutations, but the axiom is stated abstractly to capture the idea of a maximal symmetry group.) The existence of \$\Sym\$ means that we can discuss **invariances** and **transformations** that apply to absolutely everything. Any \$g \in \Sym\$ is like a “relabeling of the entire universe” that preserves structure. The group axioms ensure these transformations can be composed, have inverses, etc., just like symmetry operations in physics.

    *Rationale:* In the philosophy of science, symmetry principles are deeply linked to fundamental laws and explanatory power. By building a symmetry concept into our foundations, we allow for a **structural invariance** perspective: truths that do not depend on particular representatives but only on invariant structure. The presence of \$\Sym\$ elevates the framework from a static hierarchy to one that also considers **automorphic** or structural perspectives – analogous to how in physics the laws of nature are invariant under certain groups of transformations (Lorentz transformations, gauge symmetries, etc.). In our mathematical foundation, \$\Sym\$ can be seen as encoding the idea that even the universe of sets might have *nontrivial automorphisms* or at least that we deliberately consider transformations as a foundational ingredient. This contributes to **explanatory adequacy** by highlighting what aspects of mathematics are structural (unchanged under \$\Sym\$) and allows modeling of scenarios where different “contexts” or “frames” are related by symmetry. It also provides **structural clarity**: \$\Sym\$ helps classify mathematical objects by their symmetry orbits (see next axiom), analogous to classifying physical states by symmetry types.

\begin{enumerate}
  \item **Golden Sets (Symmetry Orbits as Contexts):** The action of \$\Sym\$ partitions \$\infty\$ into orbits, each of which will be a **Golden Set**. Formally, for any set \$x \in \infty\$, define the *orbit* of \$x\$ under \$\Sym\$ as \$\Orb(x) = {,y \in \infty : \exists g \in \Sym,\ g(x) = y,}\$. We assume:
\end{enumerate}

\begin{itemize}
  \item *(Partition into Orbits)* If \$x\$ and \$y\$ are sets in \$\infty\$, then either \$\Orb(x) = \Orb(y)\$ or \$\Orb(x) \cap \Orb(y) = \emptyset\$. Every set lies in some orbit, and orbits are pairwise disjoint.
  \item *(Orbits are Sets)* Each orbit \$\Orb(x)\$ is a set (a member of \$\infty\$). We call each such set a **Golden Set** (denoted \$G\$, \$G\_1\$, \$G\_2\$, etc. as needed). In particular, no orbit is as large as \$\infty\$ itself (the action has “multiple orbits”), so \$\Sym\$ is not transitive on the entire universe.

\end{itemize}
    *Explanation:* This axiom captures the idea that the symmetry group induces *equivalence classes* of sets, where two sets are equivalent if one can be turned into the other by some symmetry transformation. These equivalence classes (orbits) are called **Golden Sets**. The terminology “Golden” suggests these sets are special, foundational contexts – each orbit is an **internally self-contained collection** of sets that is invariant under \$\Sym\$. Formally requiring orbits to be sets (not proper classes) means each context is of strictly smaller size than the whole universe \$\infty\$. This is a strong condition ensuring that \$\Sym\$ does not just trivially act on everything as one piece, but rather there are multiple "worlds" or contexts within \$\infty\$ that \$\Sym\$ maps amongst. We can think of Golden Sets as different **domains or scales of discourse**, each closed under the symmetries of the universe relevant to that domain.

    *Rationale:* Requiring orbits to be sets provides a **hierarchy of manageable contexts** rather than one homogeneous blob. Each Golden Set is like a *universe within the universe*, but one that is smaller and self-contained. This resonates with scientific practice: we often analyze subsystems or domains (e.g. a certain energy scale in physics, or a particular organism in biology) that are approximately closed systems with respect to certain interactions. Here, a Golden Set is closed under the symmetries we consider – nothing inside it can be taken outside it by those transformations. This **invariance** gives the Golden Set an internal identity or lawfulness. It aids **explanatory adequacy** by letting us focus on one context at a time (since each Golden Set can be studied in isolation, being invariant), and yet know that other contexts exist (other Golden Sets) related perhaps by higher transformations. Philosophically, the Golden Set concept provides **structural clarity**: instead of treating the Absolute Infinite as incomprehensibly large, we recognize structured “pieces” of it (contexts) that are themselves sets and thus amenable to ordinary set-theoretic study. It bridges a gap between an ultimately large infinity and the finite/intelligible by introducing intermediate infinities that are structured and symmetric. (The name “Golden” might allude to a harmonious or optimal subset – the orbit structure yields nicely *symmetric* subuniverses, echoing how in mathematics certain invariant subsets yield deeper understanding.)

\begin{enumerate}
  \item **Hierarchical Nesting of Contexts:** The collection of Golden Sets is **nested and unbounded**. This can be expressed by two principles:
\end{enumerate}

\begin{itemize}
  \item *(Nested Universes)* If \$G\$ is a Golden Set and \$H\$ is another Golden Set with \$G \in H\$ (meaning \$G\$ is an element of \$H\$), then \$G\$ is considered a “smaller context” nested within the “larger context” \$H\$. In general, we postulate that **each Golden Set is (or is contained in) a transitive set that qualifies as a Grothendieck universe**, and Golden Sets can themselves contain other (smaller) Golden Sets as members.
  \item *(Unbounded Hierarchy)* For every Golden Set \$G\$ (except perhaps the absolutely smallest one), there exists a *strictly larger* Golden Set \$H\$ with \$G \in H\$. There is no maximal Golden Set short of \$\infty\$ itself. Thus, one gets an ascending chain \$G\_0 \in G\_1 \in G\_2 \in \cdots\$ of Golden Sets reaching “upward” towards the Absolute Infinite.

\end{itemize}
    *Explanation:* These principles ensure that the Golden Sets aren’t just a flat partition of the universe – they are arranged in a **hierarchy of levels or contexts**. The “Nested Universes” condition aligns the Golden sets with the **Grothendieck universes** from Axiom 11. In fact, we can choose to identify each Golden Set with a Grothendieck universe (or at least assume each Golden Set satisfies the Universe axioms within itself). This means each Golden Set \$G\$ is *almost* like a mini-\$\infty\$ for the sets it contains: every set \$x \in G\$ will find all its relevant supersets, power sets, etc., still inside \$G\$. So **each Golden Set is a self-contained foundational universe** for its internal members. Meanwhile, the “Unbounded Hierarchy” condition mirrors the TG axiom’s implication that there are arbitrarily large universes. Given any particular Golden Set (context), there is always a bigger context that contains it. No matter how high we climb in these nested universes, we never reach a final set that contains *all* sets – the only all-encompassing totality is \$\infty\$, which remains a proper class.

    *Rationale:* This hierarchy of Golden Sets provides a **scaffold for explanatory reduction and expansion**. In the philosophy of science, one often moves between different scales or levels of description – e.g. from particle physics (smaller context) to chemistry to biology (larger contexts), each level being to some extent self-contained but nested in a broader context. The nested Golden Set axiom captures a similar intuition: any given domain of discourse (with its own internal symmetry and laws) can be seen as a part of a larger domain with possibly richer structure. Yet, the smaller domain is not invalidated by the larger; it’s an internally coherent universe on its own. This approach is **compatible with modeling** because it allows one to work within an appropriate Golden Set (say \$G\_n\$) for a problem – knowing that \$G\_n\$ is a large enough “universe” containing all the needed mathematics for that problem (by virtue of being a Grothendieck universe), and that if necessary, one can step up to \$G\_{n+1}\$ for a wider perspective. Each Golden Set’s invariance under \$\Sym\$ (inherited from the global symmetry) means that the **laws or structures inside \$G\_n\$ are self-consistent and isolated** – an analogy to how physical laws might hold within a domain without interference from outside domains due to symmetry constraints. The unboundedness of the hierarchy speaks to an **open-ended view of science and mathematics**: there’s always a larger context or deeper theory to consider, echoing the idea that we never reach an ultimate theory of everything within the confines of set-size – the Absolute Infinite \$\infty\$ remains an aspirational ideal, not a set we fully grasp.

> **Summary of the Framework’s Philosophical Alignment:** Each axiom above not only serves a formal purpose but also aligns with goals in foundational philosophy:
>
> * *Explanatory Adequacy:* By including large infinities (TG axiom) and symmetry considerations (TOI axioms), the framework can accommodate explanations that require “stepping outside” a given system. For example, the reflection of Absolute Infinity’s properties in smaller Golden Sets and the **explanatory role of symmetries** (as Wigner and others noted, symmetries reveal deep structural explanations in physics) are built into the axioms.
> * *Structural Clarity:* The NBG axioms give a clear stratification of mathematical objects into sets and classes, preventing paradoxes and enabling talk of “the collection of all X” safely. The TG axiom further stratifies the infinite landscape by size (inaccessible cardinals), and the Golden Set axioms stratify by symmetry-invariant context. This multi-layered structure means we can always clarify “where” a mathematical or scientific statement is situated – within which context or level of the hierarchy – bringing order to the foundational picture.
> * *Compatibility with Modeling:* Scientists often isolate a system (context) to study, knowing it’s part of a larger world. Here, a Golden Set can serve as a **model context** – it’s large enough to carry out standard mathematics (being a universe in the Grothendieck sense), but also appropriately closed and symmetric, reflecting the idea that the system has its own laws (invariants under \$\Sym\$). We can model different scales or domains by different Golden Sets, and use the hierarchy axioms to **relate these models** (one context can be embedded in a larger one when needed, analogous to embedding a special-case model into a more general theory). The symmetry group \$\Sym\$ itself provides a language for mapping between scenarios: if two elements lie in the same orbit, they are in some sense *symmetric* or *analogous* across contexts, which is a powerful way to formalize the idea of analogous structures in different domains.
>
> In conclusion, this integrated axiomatic framework (NBG + TG + TOI) yields a foundation that is mathematically robust and philosophically rich. It respects classical set theory for reliability, extends it with **large-scale universes** for completeness, and innovates with **Absolute Infinity and symmetry** to incorporate a sense of overarching structure and context dynamics. Each axiom plays a role in ensuring that mathematics can serve as a foundation not just for itself but for **all of science**, by providing tools to handle unlimited size (inaccessible infinities), **invariance principles**, and a **hierarchical organization of knowledge**. The framework thus aims at a unification of Cantor’s vision of the infinite, Grothendieck’s vision of self-contained universes, and the modern view that symmetries and contexts are key to understanding complex structures in both mathematics and the natural world.

**Sources:**

\begin{itemize}
  \item Mendelson, E. *Introduction to Mathematical Logic* – (NBG class axioms and finite axiomatizability)
  \item Wikipedia: *Von Neumann–Bernays–Gödel set theory* (NBG overview, classes vs sets)
  \item Wikipedia: *Grothendieck universe* (Universe axiom and inaccessible cardinals)
  \item Karagila, A. (MathOverflow post) – Equivalence of TG axiom to many inaccessible cardinals
  \item Cantor, G. – *Correspondence* (Absolute Infinite and inconsistency); see also *Absolute Infinite* (Cantor’s reflection principle)
  \item Stanford Encyclopedia of Philosophy: *Symmetry and Symmetry Breaking* (role of symmetry in explanation).
\end{itemize}
\end{document}
